\chapter{INTRODUCTION}
\label{ch:introduction}

\section{Introduction}

Deep neural networks achieve high accuracy on image classification, speech recognition, and natural language tasks. At the same time, these models are vulnerable to adversarial examples: inputs modified by small, human-imperceptible perturbations that cause confident misclassification~\cite{szegedy2014intriguing,goodfellow2015explaining}. This vulnerability is not restricted to the digital domain. Kurakin et al.~\cite{kurakin2016adversarialphysical} demonstrated that adversarial perturbations survive printing and recapture by a camera, and similar attacks have been shown to work on physical road signs. Given the deployment of deep learning in autonomous driving, medical imaging, and security systems, understanding and addressing this vulnerability is a practical concern~\cite{yuan2019adversarial,ren2020adversarial,chakraborty2018survey,zhou2022adversarial,macas2024adversarial}.

\section{Background and Motivation}

Szegedy et al.~\cite{szegedy2014intriguing} were the first to show that well-trained neural networks can be fooled by perturbations that are nearly invisible to humans. Goodfellow et al.~\cite{goodfellow2015explaining} offered an explanation rooted in the linear behavior of neural networks in high-dimensional input spaces: a small perturbation aligned with the gradient of the loss, applied across many dimensions, accumulates into a large change in the network's output. They introduced FGSM, a single-step attack that produces adversarial examples with one gradient computation. The simplicity and effectiveness of FGSM made it clear that adversarial vulnerability is not an edge case but a systematic property of these models.

This topic was chosen to move beyond theoretical discussion and actually measure how fragile standard models are, what adversarial training costs, and how different attacks compare when evaluated under identical conditions. Implementing the attacks from scratch, rather than relying on existing libraries, was a deliberate choice. It forced a confrontation with the implementation details that papers often leave out.

\section{Problem Statement}

This thesis studies convolutional neural networks trained on MNIST and CIFAR-10 for image classification. Four questions are addressed:

\begin{enumerate}
    \item How effective are common white-box attacks (FGSM, BIM/PGD, DeepFool, Carlini--Wagner) against standard models, and how does effectiveness scale with perturbation budget?
    \item Does the picture change in the black-box setting, where the attacker has no gradient access? This is tested using HopSkipJump (decision-based) and One Pixel ($L_0$), together with a transfer-based study between the two models.
    \item Does PGD-based adversarial training yield meaningful robustness, and what is the cost in clean accuracy? Does the gain transfer across threat models, including ones the training procedure was not designed against?
    \item How do the six attacks compare in terms of attack success rate, perturbation magnitude, and computational cost?
\end{enumerate}

The scope is restricted to evasion attacks at test time in the image classification setting.

\section{Research Objectives}

Much of the existing work either remains theoretical or relies on adversarial ML libraries without investigating the underlying mechanics. To this end, the following objectives are set:

\begin{itemize}
    \item \textbf{Attack construction.} Can the six canonical attacks be reproduced from their published formulations without relying on an adversarial ML library? FGSM, BIM/PGD, DeepFool, Carlini--Wagner, HopSkipJump, and One Pixel are implemented from scratch in PyTorch against a ResNet-18 (11.17M parameters) trained on CIFAR-10 (94.1\% clean accuracy). Success is measured by driving the standard model's accuracy to 1.2\% under 20-step PGD at $\epsilon=8/255$ and by the minimum perturbation each attack requires: median $L_2$ of 0.104 (Carlini--Wagner) and 0.36 (HopSkipJump), and a single-pixel change sufficient to flip 26.0\% of correctly classified inputs at $k{=}1$.

    \item \textbf{Budget sensitivity.} How does accuracy degrade as the adversary is given more room? Each attack is evaluated across perturbation budgets ($L_\infty$ $\epsilon$ from 0 to $8/255$; $L_0$ from 1 to 5 pixels) and, for the decision-based HopSkipJump, across query budgets, to produce accuracy-versus-budget curves rather than single operating points.

    \item \textbf{Defense and its cost.} What does PGD adversarial training buy, and what does it cost? A model is trained with 7-step inner PGD ($\epsilon=8/255$, $\alpha=2/255$, random start) and measured against the standard model under an identical protocol. The target is a quantified trade-off: PGD@$8/255$ accuracy raised from 1.2\% to 45.1\% at a clean-accuracy cost of 10.2 points (94.1\% to 83.9\%), with the same comparison reported for DeepFool ($L_2$ 0.248 to 0.892), Carlini--Wagner (median $L_2$ 0.104 to 0.758), HopSkipJump (median $L_2$ 0.36 to 2.10), and One Pixel.

    \item \textbf{Transfer and threat model.} Do adversarial examples crafted on the standard model defeat the defended one? Perturbations are transferred from the standard to the robust model to measure black-box robustness, where the robust model retains 97.2--99.9\% accuracy on transferred FGSM, PGD, and Carlini--Wagner examples, complementing the white-box results and covering $L_\infty$, $L_2$, and $L_0$ threat models under one evaluation setup.

    \item \textbf{Non-parametric readout.} Can robustness be added without retraining the backbone? A modern Hopfield readout is placed over the frozen standard model's 512-dimensional features (attractors stored from the 50k training embeddings) and compared to the linear head at zero clean-accuracy cost (96.0\% both). The target is a consistent improvement on One Pixel inputs: 72.0/45.0/36.0\% versus 71.0/41.0/32.0\% at $k{=}1/3/5$.
\end{itemize}

\section{Contributions}

The contributions of this thesis are as follows:

\begin{enumerate}
    \item A controlled comparison of six adversarial attacks (spanning white-box, decision-based black-box, and transfer-based settings) is provided on the same model with the same evaluation setup. This addresses the difficulty of comparing results across papers that use different configurations.
    \item The accuracy--robustness trade-off from PGD adversarial training is reported with sufficient detail for reproducibility.
    \item An open-source framework is released containing all attack implementations, training scripts, evaluation code, and an interactive web demo.
\end{enumerate}

\section{Thesis Organization}

\textbf{Chapter 2} reviews the literature on adversarial examples, attacks, and defenses.

\textbf{Chapter 3} presents the methodology: datasets, models, attack formulations, defense setup, and evaluation metrics.

\textbf{Chapter 4} describes the implementation, including code architecture, engineering decisions, and development tools.

\textbf{Chapter 5} reports the experimental results and analysis.

\textbf{Chapter 6} gives conclusions and directions for future work.

