\chapter{LITERATURE REVIEW}
\label{ch:literature}

\section{Introduction}

This chapter reviews the existing literature on adversarial examples for deep neural networks. The theoretical foundations, the main attack methods, and the proposed defenses are covered.

\section{Theoretical Background}

\subsection{Adversarial Examples and Threat Models}

Szegedy et al.~\cite{szegedy2014intriguing} first demonstrated that neural networks with high test accuracy can be fooled by adding small, imperceptible perturbations to their inputs. Prior to this result, it was generally assumed that these networks were learning meaningful features of the data. The finding that small perturbations could cause confident misclassification suggested otherwise.

Goodfellow et al.~\cite{goodfellow2015explaining} proposed that the vulnerability arises from the linear behavior of neural networks in high-dimensional spaces. Even though the networks contain nonlinear activation functions, a perturbation that is small in each dimension can accumulate across many dimensions and produce a large change in the output. To demonstrate this, they introduced FGSM, a method that generates adversarial examples using a single gradient step. The method is simple, and the fact that it works well indicates that the vulnerability is systematic rather than incidental.

\subsection{Threat Model Taxonomy}

Adversarial attacks are typically categorized along three axes: attacker knowledge, attacker goal, and the perturbation constraint. Attacker knowledge separates the white-box setting, in which the attacker has full access to the model including its architecture, weights, and gradients, from the black-box setting, in which the attacker can only query the model and observe its outputs. Attacker goal separates untargeted attacks, which aim to cause any misclassification, from targeted attacks, which aim to produce a specific incorrect prediction. The perturbation constraint is usually an $L_p$ bound on how much the input may change: $L_\infty$ bounds the maximum change per pixel, $L_2$ bounds the Euclidean distance, and $L_0$ bounds the number of modified pixels. The six attacks studied in this thesis are placed at different corners of this space, which is what makes their comparison informative.

Yuan et al.~\cite{yuan2019adversarial} and Ren et al.~\cite{ren2020adversarial} provide thorough surveys of threat models.

\section{Related Work}

\subsection{Gradient-Based Attacks}

\subsubsection{Fast Gradient Sign Method (FGSM)}

FGSM~\cite{goodfellow2015explaining} computes the gradient of the loss with respect to the input, takes its sign, and scales it by the perturbation budget $\epsilon$, so a single forward and backward pass yields the perturbation. Because it takes only one step, FGSM does not necessarily find the strongest adversarial example within the $\epsilon$-ball, but its speed made it the first practical demonstration that high-accuracy networks are brittle to small, structured perturbations. The exact update rule used in this thesis is stated in Chapter~\ref{ch:methodology}.

\subsubsection{Basic Iterative Method (BIM) / Projected Gradient Descent (PGD)}

Kurakin et al.~\cite{kurakin2016adversarialphysical} extended FGSM by applying it iteratively with a smaller step size and clipping back into the $\epsilon$-ball after each step. Madry et al.~\cite{madry2018towards} formalized this as projected gradient descent on the inner maximization problem. The main modification was initializing from a random point within the $\epsilon$-ball, which helps the attack explore the loss landscape more effectively. PGD has since become the standard first-order attack for evaluating adversarial robustness. If a model withstands PGD, this is taken as evidence of robustness against first-order adversaries.

\subsection{Optimization-Based Attacks}

\subsubsection{DeepFool}

Moosavi-Dezfooli et al.~\cite{moosavidezfooli2016deepfool} asked a different question: what is the \emph{smallest} perturbation that changes the prediction? DeepFool answers this by iteratively linearizing the classifier and computing the distance to the nearest decision boundary. The resulting perturbations are much smaller than those of FGSM, yet they still cause misclassification. DeepFool provides geometric information that the fixed-budget attacks do not: the distance from each data point to the nearest decision boundary.

\subsubsection{Carlini-Wagner (C\&W) Attack}

Carlini and Wagner~\cite{carlini2017towards} formulated adversarial example generation as an optimization problem that jointly minimizes the perturbation norm and a term that rewards misclassification, solved with a change of variables that keeps the search in the valid pixel range. The C\&W attack demonstrated that several defenses previously claimed to be effective were in fact broken. It remains one of the strongest attacks, though it is computationally expensive compared to FGSM or PGD.

\subsubsection{Universal Adversarial Perturbations}

Moosavi-Dezfooli et al.~\cite{moosavidezfooli2017universal} showed that a single, fixed perturbation can fool a classifier on the majority of inputs. This is not an input-specific perturbation: the same noise pattern is applied to all images. The existence of such universal perturbations suggests that the vulnerability is structural rather than tied to individual inputs.

\subsection{Black-Box Attacks}

The attacks discussed so far assume white-box access: the attacker can compute gradients through the model. Realistic deployments rarely expose gradients. Commercial systems are reached through APIs that return either a probability vector or a single top-1 label. Black-box attacks address this setting, and the literature distinguishes three categories.

\subsubsection{Transfer Attacks}

Papernot et al.~\cite{papernot2017practical} showed that white-box access is not required if the attacker is willing to train a substitute. The procedure is: query the target on a small dataset, train a substitute model to imitate its outputs, generate white-box adversarial examples against the substitute, and transfer them to the target. Transfer attacks work because adversarial examples generalize across models trained on similar data. The main cost is training the substitute, after which the attacker effectively has white-box access to a proxy.

\subsubsection{Score-Based Black-Box Attacks}

Score-based attacks assume the attacker sees the full probability vector but has no gradient. Chen et al.~\cite{chen2017zoo} introduced ZOO, which estimates gradients from finite differences of the output probabilities and then applies Carlini-Wagner-style optimization. The approach works but is expensive: finite-difference gradient estimation on a $32 \times 32 \times 3$ input costs on the order of thousands of queries per image. Ilyas et al.~\cite{ilyas2018blackbox} improved this with natural evolution strategies, which estimate gradients from a smaller number of random directions and reduce the query count by roughly an order of magnitude.

\subsubsection{Decision-Based Black-Box Attacks}

Decision-based attacks make the weakest assumption: the attacker only observes the predicted label, not the probabilities. This matches what APIs such as image moderation services actually expose. Brendel et al.~\cite{brendel2018boundary} introduced the Boundary Attack, which starts from a large adversarial perturbation (for example, random noise classified as a wrong class) and walks along the decision boundary toward the clean input, shrinking the perturbation while maintaining misclassification. The attack converges to perturbations comparable in size to white-box results, but at the cost of tens of thousands of queries.

Chen, Jordan, and Wainwright~\cite{chen2020hopskipjump} proposed HopSkipJump, which is an order of magnitude more query-efficient than the Boundary Attack. The algorithm alternates between two steps: a binary search along the line between the current adversarial point and the clean input to find a point on the decision boundary, followed by a gradient-direction estimate obtained by Monte Carlo sampling of the label at nearby points. The estimated direction is used to take a step away from the boundary, and the process repeats. HopSkipJump requires only label access and converges to strong adversarial examples with a few thousand queries per image, making it one of the strongest attacks in the decision-based setting.

\subsubsection{Generative Black-Box Attacks}

A more recent line of work trains a generator network to produce adversarial perturbations directly. Xiao et al.~\cite{xiao2018advgan} introduced AdvGAN, in which a generator $G$ is trained against a discriminator $D$ and the target classifier $C$: the generator outputs a perturbation, the discriminator pushes the perturbed image to look like a clean one, and $C$ supplies the misclassification signal. Once trained, $G$ produces adversarial examples in a single forward pass. The original AdvGAN trained $G$ against a distilled white-box surrogate of $C$, which is acceptable in a black-box setting only insofar as the surrogate transfers.

Topal, Mancellari, Gillet, and Leprévost~\cite{topal2026nbugan} closed this gap with NbuGAN. The generator is trained against the black-box classifier directly, on a single image at a time, and its perturbation is produced in the low-resolution input domain of the classifier ($224 \times 224$ for ImageNet models) and then interpolated up to the original high-resolution image using the Noise Blowing-Up technique from their earlier work. The attack is targeted, requires no gradient access, and reports a 93.3\% average success rate over 100 high-resolution ImageNet images against 12 classifiers (10 CNNs and the ViT-L and ResMLP-B16 transformers), with an average attack time of $\approx 55$ seconds per image. The same paper benchmarks NbuGAN against five other black-box attacks (EA, SimBA, HopSkipJump, SignOpt, Boundary) and four white-box attacks (FGSM, BIM, PGD-$L_\infty$, PGD-$L_2$). HopSkipJump remains query-efficient for the low-resolution regime considered in this thesis, but NbuGAN is the relevant comparison once the threat model moves to native ImageNet resolution.

\subsection{Sparse and $L_0$-Bounded Attacks}

The attacks discussed so far bound the perturbation under $L_2$ or $L_\infty$. A separate line of work considers the $L_0$ constraint, which limits the \emph{number} of modified pixels rather than their magnitude. Papernot et al.~\cite{papernot2016limitations} developed the Jacobian-based Saliency Map Attack (JSMA), which iteratively modifies the pixels with the highest saliency with respect to the target class.

Su, Vargas, and Sakurai~\cite{su2019onepixel} took this to an extreme with the One Pixel Attack, which perturbs a single pixel and uses differential evolution to search for the pixel location and RGB value that cause misclassification. The attack is black-box (it requires only the probability output, not the gradient) and demonstrates that models can be fooled by changes that are localized enough to be trivially detected by a human but small enough to evade spatial averaging. Reported success rates on CIFAR-10 range from 30\% to 70\% depending on the target model, with three-pixel and five-pixel variants achieving higher rates. The interest in the One Pixel Attack is less in its practical threat and more in what it reveals about the geometry of the learned decision boundaries: a single well-placed perturbation is enough to cross them.

\section{Defense Mechanisms}

\subsection{Adversarial Training}

The most direct defense is to augment training with adversarial examples~\cite{goodfellow2015explaining,kurakin2016adversarialscale,madry2018towards}. This amounts to solving a saddle-point (min-max) problem in which the inner maximization searches for the worst-case perturbation of each input and the outer minimization updates the weights against it; the objective is stated formally in Chapter~\ref{ch:methodology}. Madry et al.~\cite{madry2018towards} showed that using PGD for the inner maximization produces substantially better robustness than using FGSM. The cost is computational: running a multi-step attack at each training iteration makes training roughly 5--10$\times$ slower. Shafahi et al.~\cite{shafahi2019adversarialfree} addressed this by reusing gradient computations, reducing the overhead of adversarial training.

\subsection{Gradient Masking and Obfuscation}

A number of defenses attempted to prevent attacks by making gradients uninformative. Athalye et al.~\cite{athalye2018obfuscated} showed that nearly all such defenses could be circumvented. They developed BPDA (backward pass differentiable approximation) and related techniques to recover useful gradients despite the defense. Many defenses that had been reported as effective turned out to provide little actual robustness.

\subsection{Other Defense Approaches}

Several other defense strategies have been explored. Input preprocessing applies transformations such as JPEG compression or denoising before classification, with the goal of removing the adversarial perturbation before it reaches the model. Certified defenses provide a formal guarantee that the prediction will not change within a given perturbation radius, at the cost of reduced accuracy. Randomized smoothing classifies by majority vote over noisy copies of the input, which yields certifiable robustness in a probabilistic sense.

Further detail is available in the surveys by Ren et al.~\cite{ren2020adversarial}, Chakraborty et al.~\cite{chakraborty2018survey}, Zhou et al.~\cite{zhou2022adversarial}, and Macas et al.~\cite{macas2024adversarial}.

\section{Research Gap}

Several gaps can be identified in the existing literature. First, most works focus on a single attack or a single defense, and a study that compares attacks as different as FGSM, PGD, DeepFool, Carlini--Wagner, and decision-based black-box methods on the same model under the same evaluation protocol is difficult to locate; because different papers use different configurations, cross-paper comparison is unreliable. Second, the accuracy--robustness trade-off is hard to assess from the literature because experimental setups vary in their $\epsilon$ values, their number of PGD steps, and their architectures, so it is often unclear whether a difference in reported numbers comes from the method or from the setup. Third, implementation details are frequently underspecified: papers report hyperparameters but omit details that matter in practice, an example being the handling of batch normalization during adversarial training, which was found here to have a substantial effect on the final model. This thesis addresses these gaps by providing a single study with a unified codebase, consistent evaluation, and sufficient implementation detail for reproducibility.

\section{Summary}

The literature indicates that adversarial vulnerability is a fundamental property of current deep learning models rather than an artifact that can be patched. Attacks range from single-step methods (FGSM) to full optimization procedures (C\&W), and the realistic black-box setting is covered by transfer, score-based, and decision-based attacks such as HopSkipJump and by sparse attacks such as One Pixel. On the defense side, adversarial training~\cite{madry2018towards} remains the most reliable approach, despite its cost in training time and clean accuracy. The remainder of this thesis builds on these foundations with the implementations and experiments developed here.

